Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-08-06 Report/Sections/appendix.tex
Update Report

Michelot \& Blackwell (2019) model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t
\end{equation}
%
where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$

@@

\begin{proof}{Proof of Theorem 1}{}
Combining Equation (\ref{eqn:OU_sol}) with Equation (\ref{eqn:CarHMM_to_OU}) gives:
%
\begin{align*}
\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{(s-1) \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
s = 1, \ldots, T-1
\end{align*}
%
If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t: (s-1) \Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:

\begin{align*}
\left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \left(\sigma^{(i)}\right)^2 \right)\\
s = 2, \ldots, T
\end{align*}
%
$X'$ follows a Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s+1})$ is consistent with that of a CarHMM with normal emission distributions.
\end{proof}

@@

\begin{proof}{Description of simulated data}{}

$\hat{Y}^*_{t,t^*}$ was simulated using the following procedure. The $k^{th}$ Fourier mode of $\hat{Y}^*_{t,t^*}$ is denoted as $\hat{Y}^{*(k)}_{t,t^*}$:

\begin{align}
	(\hat{Y}^{*(0)}_{t,0}|X^*_{t,0} = i^*) &\sim \mathcal{N} \left(0, \sigma^{*(i^*)} \right) & \nonumber \\
	%
	(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*) &\sim \mathcal{N} \left(\phi^{*(i^*)} * \hat{Y}^{*(0)}_{t,t^*-1}, \sigma^{*(i^*)} \right), & t^* = 1,2,\ldots, \lfloor Y_t/2 \rfloor \label{eqn:y_hat0} \\
	%
	\hat{Y}^{*(k)}_{t,t^*} &= a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, & k = 1,\ldots,49 \nonumber
\end{align}

\begin{align*}
    a_{t,t^*}^{(k)} &\sim  \left\{\begin{array}{lr}
	-1 & \quad w.p. \enspace 1/2 \\
	1  & \quad w.p. \enspace 1/2
	\end{array}\right. \\
	%
	(b^{(k)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
	%
	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\} \\
	{\rm{Gamma}}(250,1), & k = 1 \\
	{\rm{Gamma}}(50,1), & k = 2
	\end{array}\right. 
\end{align*}

\begin{align*}
    \hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
	%
	\hat{Y}^{*(k)}_{t,t^*}  &= -\hat{Y}^{*(100-k)}_{t,t^*}, & \qquad k = 51,\ldots,99
\end{align*}

$Y^*_{t,t^*,1:100}$ was set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:

$$Y^*_{t,t^*,1:100} = IDFT\left(\hat{Y}^*_{t,t^*}\right), \qquad t^* = 1,\ldots,\lfloor Y_t/2 \rfloor$$

$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued. $\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a two-second window. $Y^*_{t,t^*}$ is not continuous \textit{between} windows, but the jump discontinuities are not very severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+1}$ are highly correlated. See Figure \ref{fig:sim_data} for details.

From here it is straightforward to calculate both $Z^{*(1)}$ and $Z^{*(2)}$. We pick $\tilde{\omega} = 10$ periods per window, or 5 hertz. To find the distribution of $Z^{*(1)}_{t,t^*} = \mathcal{R}\left(\hat{Y}^{(0)}_{t,t^*}\right)$, use Equation (\ref{eqn:y_hat0}):
%
$$\left(Z^{*(1)}_{t,t^*} | X^*_{t,t^*} = i^* \right) = \left(\mathcal{R}\left(\hat{Y}^{*(0)}_{t,t^*}\right) | X^*_{t,t^*} = i^* \right) \sim \mathcal{N} \left(\phi^{*(i^*)} * Z^{*(1)}_{t,t^*-1}, \sigma^{*(i^*)} \right)$$
%
$Z^{*(2)}_{t,t^*}$ is the sum of gamma-distributed random variables with the same scale parameter, so the distribution of $Z^{*(2)}_{t,t^*}$ is also a gamma distribution:
%
$$Z^{*(2)}_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
%
$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
%
$$\left(Z^{*(2)}_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right)$$
\end{proof}

@@

The overall likelihood of the CarHHMM-DFT model is as follows:
%
$$\calL_{\text{CarHHMM-DFT}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N$$
%
where:
%
\begin{align*}
P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big]
\end{align*}
%
$f^{(i)}(y_t;\Theta^{(i)})$ is the emission distribution of the dive duration $y_t$ conditioned on the fact that $X_t = i$.

$\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain:
%
$$\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(i)},\Gamma^{*(i)}\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i^*)}) \mathbf{1}_N$$
%
where $P(z^*_{t,t^*}|z^*_{t,t^*-1};\Theta^{*(i)})$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$.
%
Note that $f^{(i,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-1}; \theta^{*(i,i^*)})$ is the emission distribution of $Z^*_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $Z^*_{t,t^*-1} = z^*_{t,t^*-1}$.

>>>>

\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation, which they refer to as a state-switching Ornstein-Uhlenbeck process:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t,
\end{equation}
%
where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. Unlike HMMs, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer. If the behavioral state $X_t$ is known and does not change between observations $\Big($i.e $X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1.

@@

\begin{proof}{Proof of Theorem 1}{}
Combining Equation (\ref{eqn:OU_sol}) with Equation (\ref{eqn:CarHMM_to_OU}) gives:
%
\begin{align*}
\left(Y_{s \Delta t} | X_{(s-1)\Delta t} = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y_{(s-1) \Delta t}, \enspace \left(\sigma^{(i)}\right)^2 \right),\\
s = 1, \ldots, T-1
\end{align*}
%
If $X_t$ follows a Markov chain with transitions at the observation times, then the behavioural state $X_t$ does not change between observations and we can re-index $Y_{(s-1) \Delta t} = Y'_s$ and $X_{(s-2)\Delta t} = X'_s$ for $s = 2,\ldots T$, yielding the desired result:

\begin{align*}
\left(Y'_s| X'_s = i \right) \sim \mathcal{N}\left((1-\phi^{(i)}) \mu^{(i)} + \phi^{(i)} Y'_{s-1}, \enspace \left(\sigma^{(i)}\right)^2 \right)\\
s = 2, \ldots, T
\end{align*}
%
$X'$ follows a (possibly unobserved) Markov chain, and the distribution of $(Y'_s|X'_s,Y'_{s-1})$ is consistent with that of a CarHMM with Normal emission distributions.
\end{proof}

@@

\begin{proof}{Detailed description of simulated data}{}

Assume that $X$, $Y$, and $X^*$ are known and simulated using the procedure detailed in Section \ref{sec:sim_study}. Let $DFT\{Y^*_{t,t^*}\}(k)$ be denoted as $\hat{Y}^{*(k)}_{t,t^*}$ and let $DFT\{Y^*_{t,t^*}\}$ be denoted as $\hat{Y}^*_{t,t^*}$ for shorthand.

Then, $\hat{Y}^*_{t}$ is simulated using the following procedure for every dive. First, simulate the average acceleration of the first two-second window of the dive:
$$
	(\hat{Y}^{*(0)}_{t,1}|X^*_{t,1} = i^*) \sim \mathcal{N} \left(0, \left(\sigma^{*(i^*)}\right)^2 \right)
$$
Remember that the average acceleration within a window is equal to the real part of the $0^{th}$ Fourier mode, so $\hat{Y}^{*(0)}_{t,1}$ is the average acceleration within the first window. Next, simulate the average acceleration for each subsequent two-second window:
\begin{equation}
	\Big(\hat{Y}^{*(0)}_{t,t^*}|X^*_{t,t^*} = i^*,\hat{Y}^{*(0)}_{t,t^*-100}\Big) \sim \mathcal{N} \left(\phi^{*(i^*)} \hat{Y}^{*(0)}_{t,t^*-100}, \left(\sigma^{*(i^*)}\right)^2 \right), \quad t^* = 101,201,\ldots
	\label{eqn:yhat_0}
\end{equation}
Note that $t^*$ has increments of 100 because $h=100$ when using two-second window sizes. Finally, simulate the other Fourier modes within each window:
$$
	\hat{Y}^{*(k)}_{t,t^*} = a_{t,t^*}^{(k)} i\sqrt{b^{(k)}_{t,t^*}}, \quad k = 1,\ldots,49;
$$
where $a$ and $b$ are random variables with the following distributions:
\begin{align*}
    a_{t,t^*}^{(k)} &\sim  \left\{\begin{array}{lr}
	-1 & \quad w.p. \enspace 1/2 \\
	1  & \quad w.p. \enspace 1/2
	\end{array}\right. \\
	%
	(b^{(k)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim {\rm{Gamma}}(5/k^2, 1) \\
	%
	(b^{(k)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim \left\{\begin{array}{lr}
	{\rm{Gamma}}(5/k^2, 1), \quad & k \notin \{1,2\} \\
	{\rm{Gamma}}(250,1), & k = 1 \\
	{\rm{Gamma}}(50,1), & k = 2.
	\end{array}\right. 
\end{align*}
Note that $b_{t,t^*}$ decays like $1/k^2$ to ``smooth out" the raw acceleration. For Fourier modes 50 through 99, we use the following procedure to ensure that the raw acceleration is real-valued:
$$
    \hat{Y}^{*(50)}_{t,t^*} = 0; \qquad
	\hat{Y}^{*(k)}_{t,t^*} = -\hat{Y}^{*(100-k)}_{t,t^*} \enspace \text{for} \enspace k = 51,\ldots,99.
$$
%
$\Big(Y^*_{t,t^*},\ldots,Y^*_{t,t^*+99}\Big)$ is finally set using the inverse discrete Fourier transform of $\hat{Y}^*_{t,t^*}$:
%
$$Y^*_{t,t^*+n} = IDFT\left\{\hat{Y}^*_{t,t^*}\right\}(n); \quad n = 0,\ldots,99; \quad t^* = 1,101,\ldots$$
%
%$\hat{Y}^*_{t,t^*}$ is anti-symmetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse Fourier transform is real-valued. $\hat{Y}^{*(k)}_{t,t^*}$ also decays like $1/k$ so that $Y^*_{t,t^*}$ remains continuous within a two-second window. $Y^*_{t,t^*}$ is not continuous \textit{between} windows, but the jump discontinuities are not very severe since $\hat{Y}^{*(0)}_{t,t^*}$ and $\hat{Y}^{*(0)}_{t,t^*+1}$ are highly correlated. See Figure \ref{fig:sim_data} for details.

We now prove that this construction of $Y^*$ results in the correct distributions listed in Section \ref{sec:sim_study}.

It is straightforward to calculate both $\Zone$ and $\Ztwo$ from $Y^*$. We pick $\tilde{\omega} = 10$ periods per window, or 5 Hz. To find the distribution of $\Zone_{t,t^*}$, use Equation (\ref{eqn:yhat_0}) together with the substitution $\Zone_{t,t^*} = \text{Re}\left(\hat{Y}^{(0)}_{t,t^*}\right) = \hat{Y}^{(0)}_{t,t^*}$:
%
$$\left(\Zone_{t,t^*} | X^*_{t,t^*} = i^*, \Zone_{t,t^*-100} \right) \sim \mathcal{N} \left(\phi^{*(i^*)} \Zone_{t,t^*-100}, \left(\sigma^{*(i^*)}\right)^2 \right).$$
%
This distributions is consistent with a CarHMM with Normal emission distributions and $\mu_A^{*(i^*)} = 0$. The other parameters $\sigma_A^{*(i^*)}$ and $\phi_A^{*(i^*)}$ are free, and we select $\sigma_A^{*(1)} = 0.05s$, $\phi_A^{*(1)} = 0.99$, $\sigma_A^{*(2)} = 0.1s$, and $\phi_A^{*(2)} = 0.95$.

The wiggliness $\Ztwo_{t,t^*}$ is the sum of gamma-distributed random variables with identical scale parameters, so the distribution of $\Ztwo_{t,t^*}$ is also gamma:
%
$$\Ztwo_{t,t^*} = \sum_{k=1}^{10} b^{(k)}_{t,t^*}$$
%
$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 1\right) \sim {\rm{Gamma}}\left(\alpha = \sum_{k=1}^{10} 5/k^2 = 10.10, \beta = 1.00 \right)$$
%
$$\left(\Ztwo_{t,t^*}|X^*_{t,t^*} = 2\right) \sim {\rm{Gamma}}\left(\alpha = 300 + \sum_{k=3}^{10} 5/k^2 = 305.94 , \beta = 1.00 \right).$$
%
It follows that $\mu_W^{*(1)} = 10.10$, $\sigma_W^{*(1)} = 3.18$, $\mu_W^{*(2)} = 305.94$, and $\sigma_W^{*(2)} = 17.49$.
\end{proof}

@@

The overall likelihood of the CarHHMM-DFT model is as follows:
%
$$\calL_{\text{CarHHMM-DFT}}(y,z^*;\Theta,\Theta^*,\Gamma,\Gamma^*) = \delta P(y_1,z_1^*;\Theta,\Theta^*,\Gamma^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*) \mathbf{1}_N.$$
%
In particular,
%
\begin{align*}
P(y_t,z_t^*;\Theta,\Theta^*,\Gamma^*)  = \text{diag}\Big[&f^{(1)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(1)},\Gamma^{*(1)}\right), \ldots , \\
&f^{(N)}(y_t;\Theta^{(i)})\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(N)},\Gamma^{*(N)}\right) \Big],
\end{align*}
%
where $f^{(i)}(y_t;\Theta^{(i)})$ is the emission distribution of the dive duration $Y_t$ conditioned on the fact that $X_t = i$.

The likelihood $\calL_{\text{CarHMM-DFT}}$ corresponds to the fine-scale chain and has the following expression:
%
$$\calL_{\text{CarHMM-DFT}}\left(z_t^*;\Theta^{*(i)},\Gamma^{*(i)}\right) = \delta^{*(i)} \prod_{t=2}^T \Gamma^{*(i)} P(z^*_{t,t^*}|z^*_{t,t^*-100};\Theta^{*(i^*)}) \mathbf{1}_N,$$
%
where $P(z^*_{t,t^*}|z^*_{t,t^*-100};\Theta^{*(i^*)})$ is an $N^* \times N^*$ diagonal matrix with $(i^*,i^*)^{th}$ entry equal to $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$.
%
Note that $f^{*(\cdot,i^*)}(z^*_{t,t^*}|z^*_{t,t^*-100}; \theta^{*(\cdot,i^*)})$ is the emission distribution of $\Z_{t,t^*}$ conditioned on $X^*_{t,t^*} = i^*$ and $\Z_{t,t^*-100} = z^*_{t,t^*-100}$.
